Counting And Confusion: Bayesian Rate Estimation With Multiple Populations

### 2013-01-07 many.tex
Added more detailed description of connection to Biswas

For the usual $k=1$ case \citep{Biswas2009}, the marginalized
posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{multline}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right) \\ \times \exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \\ \times \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
where $x_N$ denotes the loudness of the loudest event. Marginalizing
over $R_b$ we obtain
\begin{multline}
p_{\rm LE}\left( R_f, \theta | d \right) \propto \left(
\frac{\hat{b}(x_N,\theta)}{2(1-\hat{B}(x_N,\theta))} + R_f
\hat{f}(x_N,\theta)\right) \\ \times \frac{\sqrt{\pi}\,
  p(\theta)}{\sqrt{1-\hat{B}(x_N,\theta)}\,\sqrt{R_f}}
\exp\left(-R_f(1-\hat{F}(x_N,\theta))\right) .
\end{multline}
\begin{widetext}
This posterior has a maximum in $R_f$ at
\begin{equation}
R_f =
\frac{\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)
  +
  \sqrt{\left(\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)\right)^2
    -
    4\tilde{b}(x_N,\theta)(1-\hat{F}(x_N,\theta))\hat{f}(x_N,\theta)}}{4
  \hat{f}(x_N,\theta)(1-\hat{F}(x_N,\theta))}.
\end{equation}
where $\tilde{b}(x_N,\theta) = \hat{b}(x_N,\theta) /
(1-\hat{B}(x_N,\theta))$ and similarly for $\tilde{f}(x_N,\theta)$.\footnote{The quantity $\Lambda$ from
  \citet{Biswas2009} is related to $\hat{b}$ and $\tilde{f}$ by
  \begin{equation}
    \Lambda(x, \theta) = \frac{\tilde{f}(x,\theta)}{R_b \hat{b}(x,\theta)}.
\end{equation}}  
\end{widetext}
If $\tilde{b}(x_N,\theta) \ll 1$, we obtain the result $(1-
\hat{F}(x_N,\theta))\,R_f \approx 1/2$. This can be understood as the
statement that the rate of foreground events with ranking statistic
greater than $x_N$, $(1- \hat{F}(x_N,\theta))\,R_f$, is of order $1$,
as expected. However, $\tilde{b}(x_N,\theta) = -{\rm d}
[\ln(1-\hat{B}(x,\theta))]/{\rm d}x$ and $(1-\hat{B}(x,\theta))
\rightarrow 0$ as $x \rightarrow \infty$, so this term will be
divergent and for many reasonable examples, we will find
$\tilde{b}(x_N,\theta) >> \hat{f}(x_N,\theta)$, in which case the
posterior on $R_f$ is peaked at $0$. This issue highlights the problem
with using a loudest-event statistic with an improper prior on the
background rate $R_b$. No matter how improbable an event with $x=x_N$
is under the background distribution, it can become likely that the
event at $x_N$ is from the background distribution by taking the
background rate to be sufficiently large. Although this predicts many
more events with $x < x_N$, by using only the loudest event we do not
incorporate the information that no such events are seen. This problem
is avoided in the new framework described here, since we use all
events detected above threshold and combined rates, $R_f+R_b$,
significantly greater than the total number of observed events are
strongly disfavored.

>>>>

For the usual $k=1$ case \citep{Biswas2009}, the marginalized
posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{multline}\label{LEeq1}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right) \\ \times \exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \\ \times \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}
where $x_N$ denotes the loudness of the loudest event. In the loudest event statistic paper~\citep{Biswas2009}, the authors assume the background distribution and rate are known, which corresponds to using a delta function prior on $R_b$. They further assume a flat prior on $R_f$ and that the foreground and background distributions do not depend on any unknown free parameters. With these assumptions, the posterior on $R_f$, Eq.~\eqref{LEeq1}, is modified to
\begin{multline}
p_{\rm LE}(R_f|d) \propto  \left(R_f
\hat{f}(x_N) + R_b \hat{b}(x_N)\right) \\ \times \exp\left[-\left(
  R_f (1-\hat{F}(x_N)) + R_b (1-\hat{B}(x_N)) \right)
  \right] .
\end{multline}
Integrating over $R_f$ gives
\begin{multline}
\int_0^\infty p_{\rm LE}(R_f|d) {\rm d}R_f =  \frac{R_b \hat{b}(x_N)}{(1-\hat{F}(x_N))} {\rm e}^{-(1-\hat{B}(x_N)) R_b}\\
%\exp\left[-(1-\hat{B}(x_N))R_b\right] \\ 
\times \left(\frac{\hat{f}(x_N)}{(1-\hat{F}(x_N)) R_b \hat{b}(x_N)} + 1 \right)
\end{multline}
and so the normalised posterior is
\begin{multline}
p_{\rm LE}(R_f|d) = \frac{(1-\hat{F}(x_N))}{1+\Lambda} \left( 1+ R_f (1-\hat{F}(x_N))\Lambda \right) \\
\times \exp\left[-R_f (1-\hat{F}(x_N))\right]
\end{multline}
in which we have defined
\begin{equation}
\Lambda \equiv \frac{\hat{f}(x_N)}{(1-\hat{F}(x_N)) R_b \hat{b}(x_N)}.
\end{equation}
With the further identification $\mu \equiv R_f$ and $\hat{\epsilon} \equiv 1 - \hat{F}(x_N)$, this is Eq.~(18) of~\citep{Biswas2009} and we have shown how their parameter $\Lambda$ is related to the foreground and background distributions used here.

Returning now to Eq.~\eqref{LEeq1} and marginalizing over $R_b$, we obtain
\begin{multline}
p_{\rm LE}\left( R_f, \theta | d \right) \propto \left(
\frac{\hat{b}(x_N,\theta)}{2(1-\hat{B}(x_N,\theta))} + R_f
\hat{f}(x_N,\theta)\right) \\ \times \frac{\sqrt{\pi}\,
  p(\theta)}{\sqrt{1-\hat{B}(x_N,\theta)}\,\sqrt{R_f}}
\exp\left(-R_f(1-\hat{F}(x_N,\theta))\right) .
\end{multline}
%\begin{widetext}
%This posterior has a maximum in $R_f$ at
%\begin{equation}
%R_f =
%\frac{\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)  +
%  \sqrt{\left(\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)\right)^2  -
%    4\tilde{b}(x_N,\theta)(1-\hat{F}(x_N,\theta))\hat{f}(x_N,\theta)}}{4
%  \hat{f}(x_N,\theta)(1-\hat{F}(x_N,\theta))}.
%\end{equation}
%where $\tilde{b}(x_N,\theta) = \hat{b}(x_N,\theta) /(1-\hat{B}(x_N,\theta))$ and similarly for $\tilde{f}(x_N,\theta)$.
%\footnote{The quantity $\Lambda$ from  \citet{Biswas2009} is related to $\hat{b}$ and $\tilde{f}$ by  \begin{equation}    \Lambda(x, \theta) = \frac{\tilde{f}(x,\theta)}{R_b \hat{b}(x,\theta)}.\end{equation}}  
%\end{widetext}
This posterior has a maximum in $R_f$ at
\begin{multline}
R_f =
\frac{\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)
  + \sqrt{g(x_N,\theta)}}{4\hat{f}(x_N,\theta)(1-\hat{F}(x_N,\theta))} \\
\mbox{where }  g(x_N,\theta) =\left(\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)\right)^2
    \\ - 4\tilde{b}(x_N,\theta)(1-\hat{F}(x_N,\theta))\hat{f}(x_N,\theta)
\end{multline}
and $\tilde{b}(x_N,\theta) = \hat{b}(x_N,\theta) /(1-\hat{B}(x_N,\theta))$ and similarly for $\tilde{f}(x_N,\theta)$.

If $\tilde{b}(x_N,\theta) \ll 1$, we obtain the result $(1-
\hat{F}(x_N,\theta))\,R_f \approx 1/2$. This can be understood as the
statement that the rate of foreground events with ranking statistic
greater than $x_N$, $(1- \hat{F}(x_N,\theta))\,R_f$, is of order $1$,
as expected. However, $\tilde{b}(x_N,\theta) = -{\rm d}
[\ln(1-\hat{B}(x,\theta))]/{\rm d}x$ and $(1-\hat{B}(x,\theta))
\rightarrow 0$ as $x \rightarrow \infty$, so this term will be
divergent and for many reasonable examples, we will find
$\tilde{b}(x_N,\theta) >> \hat{f}(x_N,\theta)$, in which case the
posterior on $R_f$ is peaked at $0$. This issue highlights the problem
with using a loudest-event statistic with an improper prior on the
background rate $R_b$. No matter how improbable an event with $x=x_N$
is under the background distribution, it can become likely that the
event at $x_N$ is from the background distribution by taking the
background rate to be sufficiently large. Although this predicts many
more events with $x < x_N$, by using only the loudest event we do not
incorporate the information that no such events are seen. This problem
is avoided in the new framework described here, since we use all
events detected above threshold and combined rates, $R_f+R_b$,
significantly greater than the total number of observed events are
strongly disfavored.
